---
title: "ProbSet 8, March 6"
author: 
  - Gideon Gordon
  - <gideongordon2029@u.northwestern.edu>
date: March 6, 2026
date-format: "**[D]ue:** MMMM D, YYYY"
format: 
     pdf:
       documentclass: article
fontsize: 12pt
urlcolor: blue
number-sections: true
geometry:
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{t}
    - \floatplacement{table}{t}
    - \usepackage{flafter}
    - \usepackage[T1]{fontenc}
    - \usepackage[utf8]{inputenc}
    - \usepackage{ragged2e}
    - \usepackage{booktabs}
    - \usepackage{amsmath}
    - \usepackage{url}
---

**Due Date**: March 6, 2026  
**Submission**: https://canvas.northwestern.edu/courses/245562/assignments/1687753

---

# Problem 1

## 1a.

Compare and contrast the following transformations:
 - $\log(X)$ vs. $\sqrt{X}$
 - $\text{asinh}(X)$ vs. $\log(X + 1)$

## 1b.

Using the kidiq dataset from the slides:

```{r problem1b}
library(rosdata)
library(modelsummary)
library(plotrix)
library(kableExtra)
library(marginaleffects)
library(ggplot2)
library(sjPlot)
library(sjmisc)

data(kidiq)

# Fit three models:
# 1. Without centering: kid_score ~ mom_hs + mom_iq + mom_hs:mom_iq

non_centered_model = lm(formula = (kid_score ~ mom_hs + mom_iq + mom_hs:mom_iq),
  data = kidiq)

# 2. With centering mom_iq at its mean
kidiq$c_mom_iq <- kidiq$mom_iq - mean(kidiq$mom_iq)
centered_model = lm(formula = (kid_score ~ mom_hs + c_mom_iq + mom_hs:c_mom_iq),
                    data = kidiq)

# 3. With standardization (z-scores for mom_iq)


# Your tasks:
# 1. Fit all three models

# 2. Create a table comparing coefficients
ms.display <- modelsummary(
  list("Non-Centered" = non_centered_model, 
       "Centered" = centered_model),
  stars = TRUE, output = "kableExtra"
)
ms.display
```

I understand in theory what standardization is for. It's useful for showing us how substantive a change is based on how many standard deviations from the norm it is. But also, in practice, I don't know how to do it. So, leaving that aside... 

The interaction term in each model is the degree to which having gone to high school (HS = 1) changes the effect of mom IQ on kid IQ. 

As far as centering goes, it matters for interpretation because it makes the coefficient for `mom_hs` meaningful. Otherwise, the coefficient for `mom_hs` is the effect of going to high school when `mom_iq` is 0, which is completely nonsensical. 

---

# Problem 2

Using QOG data:

```{r problem2a}
library(rqog)
library(dplyr)
library(ggplot2)

# Load and prepare data
qog_data <- read_qog(which_data = "standard", data_type = "time-series")

democracy_data <- qog_data %>%
  filter(year == 2020) %>%
  select(
    country = cname,
    democracy = vdem_libdem,
    gdp_pc = wdi_gdpcappppcon2017,
    population = wdi_pop,
    corruption = ti_cpi
  ) %>%
  filter(!is.na(democracy), !is.na(gdp_pc), gdp_pc > 0) %>%
  mutate(
    log_gdp = log(gdp_pc),
    log_pop = log(population)
  )

# Your tasks:
# 1. Fit four different models of democracy:
#    a. Linear: democracy ~ gdp_pc
linear_dem = lm(formula = (democracy ~ gdp_pc),
                data = democracy_data)

#    b. Log-X: democracy ~ log_gdp
logX_dem = lm(formula = (democracy ~ log_gdp),
                data = democracy_data)

#    c. Log-Y: lm(log(democracy) ~ gdp_pc)  # Note: democracy is 0-1
logY_dem = lm(formula = (log(democracy) ~ gdp_pc),
                data = democracy_data)

#    d. Log-log: lm(log(democracy + 0.01) ~ log_gdp)  
# We add 0.01 because taking the log of 0 is impossible 
log_log_dem = lm(formula = (log(democracy + 0.01) ~ log_gdp),
                 data = democracy_data)



# 2. For each model:
#    a. Create residual plots
#    b. Interpret coefficients substantively
```

```{r}
res_data_linear_dem <- data.frame(fitted = fitted(linear_dem), residuals = residuals(linear_dem))
res_data_logX_dem <- data.frame(fitted = fitted(logX_dem), residuals = residuals(logX_dem))
res_data_logY_dem <- data.frame(fitted = fitted(logY_dem), residuals = residuals(logY_dem))
res_data_log_log_dem <- data.frame(fitted = fitted(log_log_dem), residuals = residuals(log_log_dem))

res_plot_linear_dem = ggplot(data = res_data_linear_dem, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Fitted Value", y = "Residual", title = "Residual Plot for Linear Model") +
  theme_minimal()

res_plot_logX_dem = ggplot(data = res_data_logX_dem, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Fitted Value", y = "Residual", title = "Residual Plot for LogX Model") +
  theme_minimal()

res_plot_logY_dem = ggplot(data = res_data_logY_dem, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Fitted Value", y = "Residual", title = "Residual Plot for LogY Model") +
  theme_minimal()

res_plot_log_log_dem = ggplot(data = res_data_log_log_dem, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Fitted Value", y = "Residual", title = "Residual Plot for Log-Log Model") +
  theme_minimal()
```

(Relied on some code from Google's AI here to get through how to set up the plots). 

For the **linear model**:
```{r}
summary(linear_dem)
res_plot_linear_dem
```
The coefficient can be interpreted as being the average difference in democracy scores between countries separated by $1 in GDP per capita. However, the values get farther from true (the fitted values are an *over*estimate) to a greater degree as the data would predict democracy scores closer to 1. 

For the *logX* model:
```{r}
summary(logX_dem)
res_plot_logX_dem
```
Here, the coefficient of GDP represents the average difference between countries that differ in GDP per capita by a factor of $e$. The residual plot shows more unpredictability as the fitted democracy scores increase (it appears there is some heteroskedasticity, based on the funnel pattern of the data), but the residuals do appear a bit smaller overall. 

For the *logY* model: 
```{r}
summary(logY_dem)
res_plot_logY_dem
```
This is tricky because taking the log of something less than 1 gets you a negative number. So all the `log(democracy)` scores will be negative, because all of the democracy scores are less than 1. The coefficient, I guess, is the factors of $e$ in democracy scores between one group of countries and another with one dollar more in GDP per capita. So a difference of one dollar in GDP per capita amounts to, on average, a difference in the log of democracy of 1.320e-05, which means I *think* multiplying the raw democracy score by $e^{1.32e-05}$. 

The residuals here are sizable, but in units of log(democracy), so that's, I *think*, an underestimate by a factor of $e$ or an overestimate by a factor of $e$.  

For the *log-log model*:
```{r}
summary(log_log_dem)
res_plot_log_log_dem

```
Okay so, the coefficient seems to mean, multiplying the `log_gdp` by $e$ corresponds to multiplying the democracy score by $e^\beta$. I THINK. I am getting confused at this point and not quite sure how to interpret a log-log plot. As far as the residuals go, we again read residuals as differences by a factor of $e$. I think. 

```{r}
# 3. Create a visualization comparing predicted vs. actual values
#    for all four models on the original scale
res_data_linear_dem = mutate(res_data_linear_dem, actual = democracy_data$democracy, type = "Linear")
res_data_logX_dem = mutate(res_data_logX_dem, actual = democracy_data$democracy, type = "LogX")
res_data_logY_dem = mutate(res_data_logY_dem, actual = democracy_data$democracy, type = "LogY", fitted = exp(fitted))
res_data_log_log_dem = mutate(res_data_log_log_dem, actual = democracy_data$democracy, type = "Log_Log", fitted = exp(fitted))
full_fitting_data_dem = rbind(res_data_linear_dem, res_data_logX_dem, res_data_logY_dem, res_data_log_log_dem)
```

```{r}
ggplot(data = full_fitting_data_dem, aes(x = fitted, y = actual, color = type)) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "Linear")) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "LogX")) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "LogY")) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "Log_Log")) +
  geom_abline(intercept=0, slope=1)
```

```{r}
# 4. Test polynomial specifications:
quadratic_model = lm(formula = (democracy ~ poly(log_gdp, 2)), data = democracy_data)
cubic_model = lm(formula = (democracy ~ poly(log_gdp, 3)), data = democracy_data)
res_data_quadratic_dem <- data.frame(fitted = fitted(quadratic_model), residuals = residuals(quadratic_model), actual = democracy_data$democracy, type = "Quadratic")
res_data_cubic_dem <- data.frame(fitted = fitted(cubic_model), residuals = residuals(cubic_model), actual = democracy_data$democracy, type = "Cubic")
full_fitting_data_dem = rbind(full_fitting_data_dem, res_data_quadratic_dem, res_data_cubic_dem)
```

```{r}
ggplot(data = full_fitting_data_dem, aes(x = fitted, y = actual, color = type)) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "Linear")) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "Quadratic")) +
  geom_smooth(data = filter(full_fitting_data_dem, type == "Cubic")) +
  geom_abline(intercept=0, slope=1)
```


**Questions:**
1. Which transformation provides the best fit? Justify using both statistical and substantive criteria.
The line I've plotted on each graph (y = x) is the line of a hypothetical model that perfectly fits the data. What we see is that of the graphs using logarithms, the `LogX` model is closest to that line. The polynomials get closer the more terms I include, but that leaves weird questions about how to interpret, for example, the cubed term. I think that the `logX` transformation offers the best fit, and also substantively allows for a straightforward interpretation, in that increasing GDP per capita by a certain amount does not directly correlate to increasing democracy by a certain amount forever. There are diminishing returns.  

# Problem 3

Using the democracy model from Problem 2:

```{r problem3a}
# Choose the best model from Problem 2
best_model <- lm(democracy ~ log_gdp, data = democracy_data)

# Create comprehensive diagnostics:
library(ggplot2)
library(patchwork)

# 1. Create all four diagnostic plots (residuals vs. fitted values, Q-Q plot of residuals, scale-location plot, and residuals vs. leverage plot) 
diagnostics_data_best <- data.frame(fitted = fitted(best_model), residuals = residuals(best_model), leverages = hatvalues(best_model), cooks = cooks.distance(best_model), country = democracy_data$country)

residuals_and_fitted = ggplot(data = diagnostics_data_best, aes(x = fitted, y = residuals)) +
  geom_text(aes(label = country), size = 3, check_overlap = TRUE) +
  labs(title = "Residuals plotted against fitted values", x = "Residuals of observations", y = "Fitted Value")

residuals_and_leverage = ggplot(data = diagnostics_data_best, aes(x = residuals, y = leverages)) +
  geom_text(aes(label = country), size = 3, check_overlap = TRUE)

# 2. Calculate and plot relevant influence statistics (Cook's distance and/or DFFITS at your discretion)
influence = ggplot(data = diagnostics_data_best, aes(x =  1:nrow(diagnostics_data_best), y = cooks)) +
  geom_text(aes(label = country), size = 3, check_overlap = TRUE) +
  labs(title = "Cook's Distance of Observations", x = "Observation Index", y = "Cook's")

# 3. Identify influential observations
# 4. Test for heteroskedasticity (Breusch-Pagan test)
```

Here's the residuals/fitted graph.

```{r}
residuals_and_fitted
```
That sure looks heteroskedastic! It's very funnel-y. 

Here's the residuals/leverage graph. I don't totally know how to interpret that. 

```{r}
residuals_and_leverage
```

And it looks like there are a few more influential observations. The most influential observations are all wealthy Gulf countries, except for Singapore. 

```{r}
influence
```

I have no idea what a Q-Q plot of residuals is. Same goes for a scale-location plot. Just... no clue. For figuring out leverage, I'm checking my old work from problem set 5 to get a set of values.

I'm also not sure what a Breusch-Pagan test is, I think maybe we went over it VERY quickly. 

**Questions:**
1. Which countries are influential observations in your model? Why?
The Gulf. Gulf countries are fairly influential relative to others, since they're high-income countries with almost no democracy (as a result of oil rents). So they're immense outliers, and they're also far from the mean x-value because most countries are not wealthy. 

2. Is there evidence of heteroskedasticity? What are the implications?
There is evidence of heteroskedasticity. I don't know how to do a Breusch-Pagan test, but I can say that just looking at the graph, it's all funnel-y. This means that, as we get to higher fitted values, a country's democracy score gets less and less predictable. In practice this is probably because of a couple of oil-rich states messing with our findings. We ought to use HC3 robust standard errors to address this problem. 

3. Are the residuals normally distributed? Why does this matter?
Honestly, hard to say; not sure how to tell (maybe that's what a Q-Q plot is for??). If residuals are not normally distributed, this might mean errors aren't normally distributed, but we sort of breezed past non-normality of errors because it's not usually very important. It may lead to some sketchy p-values and unreliable confidence intervals. 

